Cross-Loop Optimization of Arithmetic Intensity for Finite Element Local Assembly

### 2014-09-05 TACO-paper-v2/TACO-Rebuttal.txt

******************************************************************************

R3:
We answer here the 5 points of the review.
1) For what concerns execution on a GPGPU, we replicate here the same answer to 
R2.
2) We agree that profile-guided optimization could be used to improve the 
performance of kernels, although this aspect is orthogonal to most of our code
transformations, because in no way profile information could be used, for 
example, to trigger generalized loop-invariant code motion and avoidance of 
redundant computation (as explained in Section 3.2). Profile-guided 
optimization could theoretically be used for padding and data-alignment of 
arrays, provided that the compiler has sufficient machinery in place to perform 
(we quote the revised paper, Section 6) "some sort of data-flow analysis (...) 
to ensure that the extra loop iterations over the padded region do not affect 
the numerical results.". We discuss on the generality of our code 
transformation and on their integrability in a general-purpose compiler (and in 
its profile-guided optimization system) in the new Section 6 of the paper, 
which also represents novel contribution of our work.
3) TODO
4) Firedrake is *not* specifically designed for Intel architectures and 
compilers. The decision of using Intel platforms and tools is twofold: from one 
hand, we believe that Intel's is one of the most powerful vectorizing and 
optimizing compilers, so for a research like ours it was important to compare 
to the best available on the market; from the other hand, their wide 
availability represents a strong incentive for studying the problems object of 
the paper on these machines.
It is also worth noting that all general-purpose compilers are, to different 
extents, limited in optimizing for register locality and code motion (for 
instance, this is true for the Intel and GNU compilers, the latter only used 
for "local" experimentation), so our work is expected to be equally/similarly 
effective on different "standard" CPU architectures.
From the practical viewpoint, we observe that COFFEE introduces optimizations 
aimed at improving cross-loop arithmetic intensity by performing source-level 
transformations, so there are absolutely no technical issues for 
straightforwardly using our domain-specific compiler on architectures different 
than Intel. Actually, this fact is a strength of our work, because it allows 
COFFEE to be almost completely platform-independent.  The only exception 
regards intrinsics code: COFFEE is currently only capable of outputing AVX 
intrinsics, so at the moment we cannot apply vector-register tiling (the also 
called outer-product vectorization technique) to architectures different than 
x86+AVX. However, COFFEE's internal algorithms use an intermediate 
representation of intrinsics code, so extending our compiler to other 
instruction sets would be just a matter of adding a new dictionary of 
compiler-specific intrinsics.
5) We agree that expression split has many issues involved. Our cost model 
offers a simple alternative for driving expression splitting. Simplicity is the 
key to effectiveness, portability, and robustness. Probably using a graph 
partitioning approach as you rightly point out would allow us to obtain better 
performance gains, but three points are worth to be highlighted:
- The split factor is a parameter of the transformation. In our experiments we 
  have tested several split factors, which means testing as many code variants 
(i.e. as many ways of splitting an expression based on sum's associativity) 
and, as shown by the various graphs in the paper, our cost model seems to be 
capable of properly approximating such a parameter. This is of course 
encouraging.
- In the perspective of integrating expression splitting in a general-purpose 
  compiler, or generalizing this technique to other problems (as elaborated in 
the new Section 6), we feel that a simple model is preferable to keep the 
implementation complexity as low as possible.
- TODO

Other comments:
a) We elaborate on the use of loop unwinding (or unrolling) in Section 4, in 
which we have added a new paragraph (named "Loop unrolling"). This paragraph 
elaborates on "complete/partial loop unwinding" as well as other general 
aspects concerning pros and cons of applying this transformation.
b) It was copy-paste leftover, now fixed
c) Right, fixed by making the statement less strong
d) This is all clarified and explained in details in the new Section 6. The 
short answer is that we hope COFFEE can be re-used in frameworks other than 
Firedrake (e.g. the FEniCS framework that we compare to in the paper).

@@

******************************************************************************

R4:
We addressed all of the observations in your review:
1) Fixed, thanks
2) The polyhedral model is well-known among compilers people. Basically, it is 
a mathematical model for representing affine loop nests and determining in a 
systemtic way all of its possible legal transformations. Cost models are often 
used by polyhedral compilers to estimate the effectiveness of a certain 
transformation. Polyhedral optimizations often aim at improving cache locality.  
In the Related Work section, we cite one of the most relevant researches on
polyhedral compilers; note that one of the co-authors of our paper is also a 
co-author of that paper (i.e. he is an expert of polyhedral compilers).
3) The speed ups only refer to the computation of the local matrix. This is now 
clarified in Section 5.1, and also re-iterated later on in the paper. 
We used different meshes for the various experiments. There is a link to 
the code of the experiments as well as the mesh sizes employed, for each 
experiment. This is now explained in the paper, Section 5.2.1.
4)

### 2014-09-05 TACO-paper-v2/TACO-Rebuttal.txt

******************************************************************************

R3:
We answer here the 5 points of the review.
1) For what concerns execution on a GPGPU, we replicate here the same answer to 
R2.

2) We agree that profile-guided optimization could be used to improve the 
performance of kernels, although this aspect is orthogonal to most of our code
transformations, because in no way profile information could be used, for 
example, to trigger generalized loop-invariant code motion and avoidance of 
redundant computation (as explained in Section 3.2). Profile-guided 
optimization could theoretically be used for padding and data-alignment of 
arrays, provided that the compiler has sufficient machinery in place to perform 
(we quote the revised paper, Section 6) "some sort of data-flow analysis (...) 
to ensure that the extra loop iterations over the padded region do not affect 
the numerical results.". We discuss on the generality of our code 
transformation and on their integrability in a general-purpose compiler (and in 
its profile-guided optimization system) in the new Section 6 of the paper, 
which also represents novel contribution of our work.

3) TODO

4) Firedrake is *not* specifically designed for Intel architectures and 
compilers. The decision of using Intel platforms and tools is twofold: from one 
hand, we believe that Intel's is one of the most powerful vectorizing and 
optimizing compilers, so for a research like ours it was important to compare 
to the best available on the market; from the other hand, their wide 
availability represents a strong incentive for studying the problems object of 
the paper on these machines.
It is also worth noting that all general-purpose compilers are, to different 
extents, limited in optimizing for register locality and code motion (for 
instance, this is true for the Intel and GNU compilers, the latter only used 
for "local" experimentation), so our work is expected to be equally/similarly 
effective on different "standard" CPU architectures.
From the practical viewpoint, we observe that COFFEE introduces optimizations 
aimed at improving cross-loop arithmetic intensity by performing source-level 
transformations, so there are absolutely no technical issues for 
straightforwardly using our domain-specific compiler on architectures different 
than Intel. Actually, this fact is a strength of our work, because it allows 
COFFEE to be almost completely platform-independent.  The only exception 
regards intrinsics code: COFFEE is currently only capable of outputing AVX 
intrinsics, so at the moment we cannot apply vector-register tiling (the also 
called outer-product vectorization technique) to architectures different than 
x86+AVX. However, COFFEE's internal algorithms use an intermediate 
representation of intrinsics code, so extending our compiler to other 
instruction sets would be just a matter of adding a new dictionary of 
compiler-specific intrinsics.

5) We agree that expression split has many issues involved. Our cost model 
offers a simple alternative for driving expression splitting. Simplicity is the 
key to effectiveness, portability, and robustness. Probably using a graph 
partitioning approach as you rightly point out would allow us to obtain better 
performance gains, but three points are worth to be highlighted:
- The split factor is a parameter of the transformation. In our experiments we 
  have tested several split factors, which means testing as many code variants 
(i.e. as many ways of splitting an expression based on sum's associativity) 
and, as shown by the various graphs in the paper, our cost model seems to be 
capable of properly approximating such a parameter. This is of course 
encouraging.
- In the perspective of integrating expression splitting in a general-purpose 
  compiler, or generalizing this technique to other problems (as elaborated in 
the new Section 6), we feel that a simple model is preferable to keep the 
implementation complexity as low as possible.
- TODO

Other comments:
a) We elaborate on the use of loop unwinding (or unrolling) in Section 4, in 
which we have added a new paragraph (named "Loop unrolling"). This paragraph 
elaborates on "complete/partial loop unwinding" as well as other general 
aspects concerning pros and cons of applying this transformation.

b) It was copy-paste leftover, now fixed

c) Right, fixed by making the statement less strong

d) This is all clarified and explained in details in the new Section 6. The 
short answer is that we hope COFFEE can be re-used in frameworks other than 
Firedrake (e.g. the FEniCS framework that we compare to in the paper).

@@

******************************************************************************

R4:
We addressed all of the observations in your review:
1) Fixed, thanks

2) The polyhedral model is well-known among compilers people. Basically, it is 
a mathematical model for representing affine loop nests and determining in a 
systemtic way all of its possible legal transformations. Cost models are often 
used by polyhedral compilers to estimate the effectiveness of a certain 
transformation. Polyhedral optimizations often aim at improving cache locality.  
In the Related Work section, we cite one of the most relevant researches on
polyhedral compilers; note that one of the co-authors of our paper is also a 
co-author of that paper (i.e. he is an expert of polyhedral compilers).

3 and 4) The speed ups only refer to the computation of the local element 
matrix.  This is now clarified in Section 5.1, and also re-iterated later on in 
the paper. We used different meshes for the various experiments. In Section 
5.2.1 we provide a link to a repository in which the code of the experiments as 
well as the mesh sizes employed can be found. We note that for these 
experiments the mesh size was set such that the working set of the computation 
could at least fit the last level cache (L3 on the Sandy Bridge architecture). 
The goal of these experiments (Section 5.2 in the revised paper) was only to 
measure speed ups achieved by optimizing local assembly kernels, so there was 
no point in running over big meshes (i.e. meshes that would not fit the last 
level cache), which would have only increased the completion time of our test 
suite (which is already order of hours).
We however completely agree that reporting execution times and speed ups of the 
overall finite element calculation is of fundamental importance. Therefore, 
following your suggestion, we have added a new section to the paper, Section 
5.3, which is about a full-application study concerning a linear elasticity 
equation.  This means that the performance evaluation section is now structured 
in two parts: Section 5.2, about local assembly speed ups; Section 5.3, about a 
real application study. We think this is the best structuring to let readers 
understand the effectiveness of our transformations "at the level of loop 
nests" as well as their impact on the whole application. In the new section, we 
provide the numbers you suggested: we show execution times of assembly and 
solve, we comment on the achieved speed ups, and we highlight the cases in 
which matrix insertion is negligible because of the high computational cost of 
local assembly. For this application, we use a realistic mesh composed of 
196608 cells (as reported in the paper). We also provide a link to the code of 
the application.
